Simple repression draft

### 2020-05-27 doc/section_04_bayesian_inference.tex
RP on inference
Made it up to and through Figure 3

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig2:constit_cartoons}. these models have Fano factors $\nu\le 1$
meaning they are underdisperse relative to Poisson. We will also not explicitly
consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

@@

\mrm{
We alert the reader not to confuse a tightly constrained posterior distribution
with a good generative model. In other words, not because the single parameter
for our Poisson likelihood can only take a small range of values constrained by
the data, this implies that such Poisson distribution characterizes the
underlying cell-to-cell variability we observe in the data. To systematically
check the validity of the generative model we can make use of \textit{posterior
predictive checks}. These consists on generating synthetic data using our
generative model for many plausible values of our parameter. For the case of our
\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
samples out of a Poisson distribution with such $\lambda$ value. We then repeat
this procedure 1000 times. Clearly if those data appear quite different, the
model has a problem. Put another way, if we suppose the generative model is
true, then the synthetic datasets we generate should resemble the actual
observed data, and if not, it suggests the model is missing important features.}

\mrm{
To compare so many samples with the actual observed data, we can use empirical
cumulative distribution functions (ECDF) of the distribution quantiles. This 
representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
example, the median for each possible mRNA count for the Poisson distribution
is shown as a dark green line, while the lighter green contains 95\% of the
posterior predictive samples. This way of representing the posterior predictive
checks gives us a sense of the range of data we might consider plausible, under
the assumption that the model is true. In this case it is quite obvious that the
observed data, plotted in orange, could not plausibly come from this Poisson
generative model. We will cover the other model shown in
Figure~\ref{fig:constit_post_full}(B) next.
}

@@

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood --the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one to two.
\mrm{
Details on the arguments behind our prior distribution selection are left for
Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2) \\
\ln b \sim \text{Normal}(0.5, 1) \\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
Note that we specify a distribution over the logarithm of our parameters $k_i$
and $b$. This is commonly the case for parameters for which our previous 
knowledge based on our domain expertise spans several orders of magnitude.
}

>>>>

It will be instructive, however, to first consider the Poisson promoter, model 1
in Figure~\ref{fig2:constit_cartoons}. As we alluded to earlier, since the
Poisson distribution has a Fano factor $\nu$ strictly equal to 1, and all of
the observed data has Fano factor $\nu>1$, we might already suspect that this
model is incapable of fitting the data. We will verify that this is in fact the
case. Using the same argument we can immediately rule out models 2 and 3 from 
Figure~\ref{fig2:constit_cartoons}.  These models have Fano factors $\nu\le 1$
meaning they are underdispersed relative to the Poisson distribution. We will also not explicitly
consider model 4 from~\fig{fig2:constit_cartoons} since it was already thoroughly
analyzed in~\cite{Razo-Mejia2020}, and since model 5 can be viewed as a special
case of it.

@@

We alert the reader not to confuse a tightly constrained posterior distribution
with a good generative model. In other words, not because the single parameter
for our Poisson likelihood can only take a small range of values constrained by
the data, this implies that such Poisson distribution characterizes the
underlying cell-to-cell variability we observe in the data. To systematically
check the validity of the generative model we can make use of \textit{posterior
predictive checks}. This consists in generating synthetic data using our
generative model for many plausible values of our parameter. For the case of our
\textit{lacUV5} promoter this means that we sample a value for $\lambda$ out of
the posterior distribution (Eq.~\ref{eq:posterior_poisson}), and generate 2648
samples drawn from a Poisson distribution with  $\lambda$ chosen appropriately. We then repeat
this procedure 1000 times. Clearly if those data appear quite different, the
model has a problem. Put another way, if we suppose the generative model is
true, then the synthetic datasets we generate should resemble the actual
observed data, and if not, it suggests the model is missing important features.

To compare so many samples with the actual observed data, we can use empirical
cumulative distribution functions (ECDF) of the distribution quantiles. This 
representation is shown in Figure~\ref{fig:constit_post_full}(B). In this 
example, the median for each possible mRNA count for the Poisson distribution
is shown as a dark green line, while the lighter green contains 95\% of the
posterior predictive samples. This way of representing the posterior predictive
checks gives us a sense of the range of data we might consider plausible, under
the assumption that the model is true. In this case it is quite obvious that the
observed data, plotted in orange, could not plausibly come from this Poisson
generative model. We will cover the other model shown in
Figure~\ref{fig:constit_post_full}(B) next.

@@

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use in Bayes theorem. Again
denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots, m_N\}$, here
Bayes theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood --the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom}--  so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one to two.
Details on the arguments behind our prior distribution selection are left for
Appendix~\ref{sec:bayesian}. Our full generative model can be specified as
\begin{equation}
\begin{split}
\ln k_i \sim \text{Normal}(-0.5, 2) \\
\ln b \sim \text{Normal}(0.5, 1) \\
m \sim \text{NBinom}(k_i, b).
\end{split}
\end{equation}
Note that we specify a distribution over the logarithm of our parameters $k_i$
and $b$. This is commonly the case for parameters for which our previous 
knowledge based on our domain expertise spans several orders of magnitude.
